---
title: "Research of Board Games"
subtitle: "NSC-R Tidy Tuesday February 2022"
author: Asier Moreva
date: "2022-02-08"
format:
    html 
---

## Introduction

The dataset for this Tidy Tuesday is about board games! Asier Moneva explores the following research questions: "What are the top 3 best-selling board games by game type?" and "How do board game sales relate to online user ratings?"[@moneva_nsc-r_2022]

## Load packages

```{r}
#| warning: false
#| echo: true
library(dplyr)
library(ggplot2)
```

Today's data are relational, which means there are multiple datasets (`rating`, `details`) that can be linked through some identifier or `id` variable or 'key'.

## Import data

```{r}
#| warning: false
#| echo: true
ratings <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-25/ratings.csv")
```

```{r}
#| warning: false
#| echo: true
details <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-25/details.csv")
```

## Explore data

So let's have a quick look at the data.

```{r}
#| warning: false
#| echo: true
glimpse(ratings)
glimpse(details)
```

You see that the number of observations do not match!

Explore whether both datasets have identical `id`

```{r}
#| warning: false
#| echo: true
identical(details$id, ratings$id)
```

Count how many `id` are missing.

```{r}
#| warning: false
#| echo: true
table(ratings$id %in% details$id)
```

Find which `id` are present in both datasets. Results not show here because of too long output.

```{r}
#| warning: false
#| echo: true
#| output: false
intersect(details$id, ratings$id)
```

Find which `id` are absent in the dataset with the least observations

```{r}
#| warning: false
#| echo: true
setdiff(ratings$id, details$id)
```

Merge data

```{r}
#| warning: false
#| echo: true
df <- inner_join(ratings, details, by = "id")
```

Explore data

```{r}
#| warning: false
#| echo: true
glimpse(df)
```

Looks like there are some duplicated variables!

Not really

```{r}
#| warning: false
#| echo: true
identical(df$year, df$yearpublished)
```

What is the difference?

```{r}
#| warning: false
#| echo: true
which(df$year != df$yearpublished)
```

Ah, I see!

```{r}
#| warning: false
#| echo: true
df$year[21623]
df$yearpublished[21623]
```

Nope. Very different. Can these be row numbers (e.g. spreadsheet)?

```{r}
#| warning: false
#| echo: true
identical(df$num.x, df$num.y)
```

Almost, but nope.

```{r}
#| warning: false
#| echo: true
identical(df$name, df$primary)
which(df$name != df$primary)
```

Looks like there is some sort of syntax issue in `primary`

```{r}
#| warning: false
#| echo: true
df$name[19925]
df$primary[19925]
```

Bingo!

```{r}
#| warning: false
#| echo: true
identical(df$playingtime, df$maxplaytime)
```

Let's drop the duplicated variable

```{r}
#| warning: false
#| echo: true
df <- df %>% 
  select(
    - playingtime,
    - primary
  )
```

Looks like we are good now! Imagine the following:

-   We want to buy a board game for a friend, but we don't know which one to get.

-   We know our friend likes to play alone with his girlfriend sometimes, but some Other times he likes to invite friends over to play.

-   What are the top 3 best-selling board games by game type?

-   How do board game sales relate to online user ratings?

-   What are the top 3 best-selling board games by game type?

```{r}
#| warning: false
#| echo: true

df <- df |> 
  # Create new variable to define type of game based on number of players
  mutate(minplayers_fct = case_when(
    minplayers < 2 ~ "single player",
    minplayers == 2 ~ "two players",
    minplayers > 2 ~ "multiplayer"
  ))
```

```{r}
#| warning: false
#| echo: true
df |> 
  # Select only relevant variables
  select(
    name,
    minplayers_fct,
    owned
  ) |> 
  group_by(minplayers_fct) |> 
  slice_max(owned, n = 3) |> 
  knitr::kable()
```

We have identified the best-sellers for each category, but best-sellers might be a product of marketing rather than an indicator of good quality. It might be the case that players buy the product because of good marketing but then find the game boring, too complicated, etc.

How do board game sales relate to online user ratings?

## Plot data

```{r}
#| warning: false
#| echo: true
#| label: patterns_players
#| fig-cap: "Relation board game sales and online user ratings"
#| cap-location: margin
#| fig-width: 6
#| fig-height: 4
df %>% 
  ggplot(mapping = aes(
    x = average,
    y = owned,
    # Examine if there are clear patterns by number of players
    color = minplayers_fct
  )) +
  geom_point(alpha = .5) +
  # There seems to be three outliers that are best-sellers and also have high 
  # ratings
  ggrepel::geom_text_repel(
    data = df |> filter(owned > 150000),
    mapping = aes(label = name),
    color = "black"
    ) +
  scale_x_continuous(limits = c(0, 10)) +
  scale_color_viridis_d() +
  labs(
    title = "How do board game sales relate to online user ratings?",
    x = "average user rating",
    y = "number of\ncopies owned",
    color = "Category"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(
    angle = 0,
    vjust = .5
  ))
```

## References
